---
title: "Data formating from the Řimov reservoir"
format:
  html:
    toc: true
    toc-depth: 4
    toc-title: Contents
    toc-location: left-body
    embed-resources: true
    self-contained-math: true
date: today
date-format: long
editor: visual
author: Samuel Dijoux
execute:
  echo: true
  warning: false
  message: false
---

Load the required libraries

```{r, warning=FALSE}

library(datetimeutils) # convertion of date/time between R Excel
#library(dplyr)
library(readxl) # to read excel file
library(tidyverse)
```

## 1) Formating the environmental and phytoplantonic dataset

The Rimov dataset is composed of 5 variables related to the sampling period, 28 variables related to the Morpho-Functional Groups of phytoplankton, 6 variables size-class related to zooplankton (4 for cladocerans and 2 for copepods), and 18 physico-chemical variables related to the environmental data.

```{r}
Rimov <- readxl::read_excel("Data/TS_Rimov_data.xlsx", col_names=TRUE, na='NA', col_types=c("date", rep("numeric", 54)) );

# Record the date month of sampling and add the year phase (qualified below)
Rimov <- Rimov %>% #filter(Year > 1999 & Year < 2021) %>% 
  mutate(Month = as.numeric(format(Date, "%m")),
         Phase = rep(0, length = dim(Rimov)[1])) %>%
  relocate("Month", "Phase", .before = Sampling)

# The distinct Periods of the Reservoir (source: Znchor et al. (2020) Sci. Tot. Env)
Rimov[which(Rimov$Year < 1990),]$Phase <- 1 # for years within the period 1983-1989
Rimov[which(Rimov$Year > 1989),]$Phase <- 2 # for years within the period  1990-1998
Rimov[which(Rimov$Year > 1999),]$Phase <- 3 # for years within the period  2000-2014
Rimov[which(Rimov$Year == 1999),]$Phase <- NA # Uncategorized year (1999)

Rimov <- Rimov %>% drop_na()
```

For simplicity we separate each dataset and focus only on the variables related to the environment and the phytoplankton MGFs.

```{r}
Sampling <- Rimov[,c(1:5)] %>% as.data.frame()
Env <- Rimov[,c(40:57)] %>% as.data.frame()
```

We restrain the number of environmental variables to kept those of most interest.

```{r}
Env <- Env %>% select(DO, pH, Conduct, DOC, TP, TN, DIN, "Surf-alti", Temp) 
```

We also calculate the total biovolume of phytoplankton sampled at a given time.

```{r}
Phyto <- Rimov %>% 
  mutate(Phyto_BV = apply(select(.,starts_with("MFG")), 1, sum)) %>%
  select(starts_with("MFG"), Phyto_BV)
```

We then calculate the relative proportion of each MFGs at any given time. Due to the amount of MFG variables, we filter the dataset to only keep MFGs which represent $2.5 \%$ of the phytoplankton biovolume in average over the time period.

```{r}
Phyto_prop <- Phyto %>%
  mutate_at(.vars=vars(matches("MFG")),~unlist(list((./Phyto_BV) ))) %>%
  select_if(.vars=vars(matches("MFG")), ~unlist(list( mean(., na.rm=T) > 0.025)) ) %>%
  select(-Phyto_BV)
```

We also restrain the `Phyto` dataset with the selected variables.

```{r}
Phyto <- Phyto %>% select(colnames(Phyto_prop) )
```

## 2) Normalization of all variables with

The normalization of each environmental and MFGs variable $x$ is performed by applying the formula: $$ x - \frac{\mu_x}{\sigma_x}$$ with $\mu_x$ the zero mean value of $x$ and $\sigma_x$ the standard deviation of 1.

```{r}
normalize <- function(x, ...) {(x - mean(x, na.rm=T))/sd(x, na.rm=T)} 
```

The normalization function applied to each dataset:

```{r}
Env_std <- Env %>% mutate_all(.funs = normalize)
Phyto_std <- Phyto %>% mutate_all(.funs = normalize)
Phyto_prop_std <- Phyto_prop %>% mutate_all(.funs = normalize)
```

We reconstitute the Rimov dataset together, with the normalized variables.

```{r}
Rimov_std <- cbind.data.frame(Sampling, Phyto_std, Env_std)
```

## Saving the *.RData* of all datasets and functions

```{r}
save.image("Data/Rimov.datasets.RData")
```
