---
title: "02.Rimov_EDM_analysis"
format:
  html:
    toc: true
    toc-depth: 4
    toc-title: Contents
    toc-location: left-body
    embed-resources: true
    self-contained-math: true
date: today
date-format: long
editor: visual
author: Samuel Dijoux
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This script is dedicated to the empirical dynamical modelling analyses performed on time series of different morpho-functional groups of phytoplanktonic communities sampled in the Řimov reservoir (Czech Republic) between 1997 and 2021. We aim to address the question of how the different morpho-functional phytoplanktonic groups interact with one another regarding the fluctuating physical and chemical changes in the Rimov reservoir across the time period. 

This document follows similar steps performed by *Liu and Gaines (2022). Environmental context dependency in species interactions* (PNAS), available here:

-   Paper <https://www.pnas.org/doi/abs/10.1073/pnas.2118539119?af=R>

-   Analysis <https://github.com/owenrliu/env_context_dependency>.


## Set up

Here is the state of the system with the different packages used. Load the required packages (if these are not installed on your system, use `install.packages()`)

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(tidyr)
library(knitr)
library(kableExtra)
library(purrr)
library(dplyr)
library(tibble)
library(readr)
library(stringr)
library(lubridate)
library(forcats)
library(fields)

#devtools::install_github("ha0ye/rEDM")
library(rEDM)

library(ncdf4)
library(quantreg)
library(RANN)
library(ggsci)
library(here)

library(ggplot2)
library(plot3D)
library(igraph)
library(gridExtra)

library(RColorBrewer)
```

```{r, echo=FALSE}
print(sessionInfo())
```

### Plot theme

The following chunk defines the plot theme used across the analyses

```{r}
plot_theme <-   theme_minimal()+
  theme(text=element_text(family="sans",size=12,color="black"),
        legend.text = element_text(size=14),
        axis.title=element_text(family="sans",size=14),
        axis.text=element_text(family="sans",size=8),
        strip.background = element_rect(colour="black"),
        panel.border = element_rect(color="black",fill=NA))

theme_set(plot_theme)
```

# Description of the Rimov reservoir and studied Variables

Brief state of art of the Rimov reservoir (missin for now, to be added).

```{r, echo=FALSE}
load("Data/Rimov.datasets.RData")
```

```{r, echo=FALSE}
key.table <- tibble(dataset=c(rep("MFG Biovolume", 11),rep("Environment",9)),
                  short=c( names(Phyto), names(Env)),
                  long = c( paste0("MFG ", substr(colnames(Phyto_std), 5, 6)),
                           "Dissolved oxygen", "pH", "Conductometry", "Dissolved organic carbon",
                           "Total Phosphorus concentration", "Total Nitrogen concentration",
                           "Dissolved inorganic nitrogen", "Altitude of the water surface (m.a.s.l)", "Temperature"),
                  col = c( brewer.pal(11, 'RdYlBu'), brewer.pal(9, 'Paired')))
key.table$col[6] <- rgb(0,0,0)
```

```{r, warning=FALSE, echo=FALSE}
pd <- position_dodge(width=2)
variables <- c( names(Phyto), names(Env))

ts_list <- list()
for(i in 1:length(variables)){
  yname <- ifelse(i<=11,
                  paste0("Normalized Biovolume of MFG ", substr(variables[i], 5, 6)),
                  paste0("Normalized Index Value of ", variables[i]))
  
  data <- Rimov_std %>% select(Date, variables[i])
  colnames(data)[2] <- 'value'
  
  pp <- data %>%
    ggplot(aes(x=Date, y = value))+
    geom_hline(yintercept=0, lty=2)+
    geom_line(lwd=1, alpha=0.8, size=0.3, color = key.table[i,]$col)+
    xlab("Year")+ylab(yname)
  
  ts_list[[i]] = pp
}
```

Our dataset is composed of 11 morpho-functional groups (MFGs) of phytoplankton and up to 9 environmental variables, all processed in the same time period. The 11 selected MFGs were selected as they contributed to at least $2.5 \%$ to the total phytoplanktonic biovolume in average over the time period.

To perform empirical dynamic models, it is best to create normalized time series in order to not distort the state-space reconstruction due to differences in magnitudes between variables. For this, we thus normalized our data using a zero mean and standard deviation of 1. See the script `01. Rimov_data_format` for more details and codes.  

```{r}
tibble(Variable=colnames(Phyto_std),
       Description = c(paste0("Normalized biomass of MFG ", substr(colnames(Phyto_std), 5,  6) ) )) %>% knitr::kable()
```

```{r}
tibble(Variable=colnames(Env_std),
       Description = paste0("Normalized ", c(
  "Dissolved oxygen", "pH", "Conductometry", "Dissolved organic carbon",
  "Total Phosphorus concentration", "Total Nitrogen concentration",
  "Dissolved inorganic nitrogen", "Altitude of the water surface (m.a.s.l)", "Temperature" ) )) %>% knitr::kable()
```



-     Morpho functional groups

```{r, echo=FALSE}
for(i in 1:11){
  print(ts_list[[i]])
}
```

-     Environmental variables

```{r, echo=FALSE}
for(i in 12:20){
  print(ts_list[[i]])
}
```

# Univariate analysis

Our goal by using the EDM technique is to reconstruct the behaviour of dynamic state of the system from our time series.
For this we need to check that our time series are valid

We need to verify that each of our time series are properly embedded
and are non-linear
For this, we will use lags of each single time series
Instead of using a complete set of state variables to represent the system, 

Each time series can

Multiple preliminary steps are required before being able to

Simplex projection


Before testing the different causal interactions between variables, we need to run a few preliminary steps in order to reconstruct 

We need evidence that the variables are

-   the optimal lag (or embedding dimension) which unfold the best the time series

-   their ability to self predict, i.e by detecting whether the time series is stochastic or not (notably by testing whether the variables have a chaotic pattern or not)

-   the variable is non linear

## Simplex, to identify Optimal embedding dimensions

we use Nearest neighbor forecasting method (Simplex) 

We start by identifying the optimal embedded dimension $E$ that best unfold the dynamics of each variables. This is assessed by selecting the embedded dimension for which we observe the highest prediction skill $\rho$.

```{r, echo=FALSE}
Rimov.Edim.list <- list(); j=1
for(i in 6:dim(Rimov_std)[2]){
  out <- EmbedDimension(dataFrame=Rimov_std, 
                        lib=c(1, floor(dim(Rimov_std)[1]/2) ),
                        pred=c(floor(dim(Rimov_std)[1]/2)+1, dim(Rimov_std)[1]),
                        columns = colnames(Rimov_std)[i], target = colnames(Rimov_std)[i], showPlot=F) %>% 
    mutate(ID = key.table$long[match(colnames(Rimov_std)[i],key.table$short)])
  Rimov.Edim.list[[j]] <- out; j <- j+1
  
}
rm(j, out)

# We saves the best dimension
Rimov.Edim.best <- sapply(Rimov.Edim.list, function(df){ df[which.max(df$rho),]$E })
maxE <- cbind.data.frame(E=Rimov.Edim.best, ID=key.table$long)
```

-     Morpho-functional groups

```{r, echo=FALSE}
p1 <- bind_rows(Rimov.Edim.list[1:11]) %>%
  ggplot( aes(E, rho, colour = ID) )+
  geom_line(size=2)+
  facet_wrap(~ID, nrow=3, ncol=4, scales="free_y")+
  labs(x="Embedding Dimension (E)",y=expression(paste("Prediction Skill, ",rho)))+
  scale_x_continuous(breaks=seq(0,12,by=2))+
  guides(color=F)

p1+geom_vline( aes(xintercept=E), size=1, linetype=2, maxE[c(1:11),])
```

-     Environmental variables

```{r, echo=FALSE}
p2 <- bind_rows(Rimov.Edim.list[12:20]) %>%
  ggplot( aes(E, rho, colour = ID) )+
  geom_line(size=2)+
  facet_wrap(~ID, nrow=3, ncol=3, scales="free_y")+
  labs(x="Embedding Dimension (E)",y=expression(paste("Prediction Skill, ",rho)))+
  scale_x_continuous(breaks=seq(0,12,by=2))+
  guides(color=F)

p2+geom_vline( aes(xintercept=E), size=1, linetype=2, maxE[c(12:20),])
```

## Prediction decay

how prediction skill changes as we increase the "time to prediction" argument
For non linear systems, we expect t
Demonstration 
short-term prediction 

```{r, echo=FALSE}
Rimov.Tp.list <- list(); j=1
for(i in 6:dim(Rimov_std)[2]){
  out <- PredictInterval(dataFrame=Rimov_std, 
                        lib=c(1, floor(dim(Rimov_std)[1]/2) ),
                        pred=c(floor(dim(Rimov_std)[1]/2)+1, dim(Rimov_std)[1]),
                        columns = colnames(Rimov_std)[i], target = colnames(Rimov_std)[i],
                        E = maxE$E[j], showPlot=F) %>% 
    mutate(ID = key.table$long[match(colnames(Rimov_std)[i],key.table$short)])
  Rimov.Tp.list[[j]] <- out; j <- j+1
  
}
rm(j,out)
```

-     Morpho-functional groups

```{r, echo=FALSE}
(q1 <- bind_rows(Rimov.Tp.list[1:11]) %>%
  ggplot( aes(Tp, rho, colour = ID) )+
  geom_line(size=2)+
  facet_wrap(~ID, nrow=3, ncol=4, scales="free_y")+
  labs(x="Time to prediction (Tp)",y=expression(paste("Prediction Skill ",rho)))+
  scale_x_continuous(breaks=seq(0,12,by=2))+
  guides(color=F))
```

-     Environmental variables

```{r, echo=FALSE}
(q2 <- bind_rows(Rimov.Tp.list[12:20]) %>%
  ggplot( aes(Tp, rho, colour = ID) )+
  geom_line(size=2)+
  facet_wrap(~ID, nrow=3, ncol=3, scales="free_y")+
  labs(x="Time to prediction (Tp)",y=expression(paste("Prediction Skill ",rho)))+
  scale_x_continuous(breaks=seq(0,12,by=2))+
  guides(color=F))
```


## S-maps, to identify nonlinear time series

We now analyse the time series to test their nonlinearity using S-maps (short for "sequentially weighted global linear maps"). This method fits local linear maps to describe the dynamics. The nonlinear localisation parameter $\theta$ determines the degree to which individual points are weighted when fitting the local linear map. All points are equally weighed for $\theta = 0$, meaning that the local linear map is identical for all points in the reconstructed state-space, and that the S-map will identical to a global linear map (obtained for an autoregressive model). Oppositely, nearby points receive larger weights in the state-space for $\theta > 0$, and the local linear maps will differ in state-space to accommodate nonlinear behavior.

Thus greater predictive skill values $\rho$ observed
-   for $\theta = 0$ suggest that the S-map will be identical to a global linear map, i.e time series may be sampled from autoregressive red noise
-   for $\theta > 0$ suggest nonlinear dynamics. Better forecasts are achieved when the local linear map can change depending on the location map $\theta$.

Following, we build the S-maps for the different time series using the optimal embedding dimensions previously analysed.

```{r, echo=FALSE}
Rimov.Smaps.list <- list(); j=1
for(i in 6:dim(Rimov_std)[2]){
  out <- PredictNonlinear(dataFrame=Rimov_std, 
                        lib=c(1, floor(dim(Rimov_std)[1]/2) ),
                        pred=c(floor(dim(Rimov_std)[1]/2)+1, dim(Rimov_std)[1]),
                        columns = colnames(Rimov_std)[i], target = colnames(Rimov_std)[i],
                        E = maxE$E[j], showPlot=F) %>% 
    mutate(ID = key.table$long[match(colnames(Rimov_std)[i],key.table$short)])
  Rimov.Smaps.list[[j]] <- out; j <- j+1
}
rm(j,out)

# We saves the best dimension
Rimov.Theta.best <- sapply(Rimov.Smaps.list, function(df){ df[which.max(df$rho),]$Theta })
maxTh <- cbind.data.frame(Theta=Rimov.Theta.best, ID=key.table$long)
```

-     Morpho-functional groups

```{r, echo=FALSE}
r1 <- bind_rows(Rimov.Smaps.list[1:11]) %>%
  ggplot( aes(Theta, rho, colour = ID) )+
  geom_line(size=2)+
  facet_wrap(~ID, nrow=3, ncol=4, scales="free_y")+
  labs(x=expression(paste("S-map localisation ", theta)),y=expression(paste("Prediction Skill  ",rho)))+
  scale_x_continuous(breaks=seq(0,12,by=2))+
  guides(color=F)

r1+geom_vline( aes(xintercept=Theta), size=1, linetype=2, maxTh[c(1:11),])
```

For all MFGs, best prediction skill is improved for positive $\theta$ across all outputs, suggesting nonlinear dynamics for these variables. Declines in forecast skill past the optimal $\theta$ (in dashed lines) indicate that local linear maps overfit to insufficient neighbors.

-     Environmental variables

```{r, echo=FALSE}
r2 <- bind_rows(Rimov.Smaps.list[12:20]) %>%
  ggplot( aes(Theta, rho, colour = ID) )+
  geom_line(size=2)+
  facet_wrap(~ID, nrow=3, ncol=3, scales="free_y")+
  labs(x=expression(paste("S-map localisation ", theta)),y=expression(paste("Prediction Skill ",rho)))+
  scale_x_continuous(breaks=seq(0,12,by=2))+
  guides(color=F)

r2+geom_vline( aes(xintercept=Theta), size=1, linetype=2, maxTh[c(12:20),])
```

5 of our 9 selected drivers seem to be characterized by linear dynamics, which higher predictive skill observed at $\theta = 0$ for Altitude of the water surface, DIN, DOC, TN and TP.

The remaining 4 drivers (Conductometry, DO, pH and Temperature) seems to be characterized by nonlinear dynamics.

## Overall summary

In overall, 

the Simplex, prediction horizon and S-maps results suggest our approach is valid for:
-

but only for Conductometry, DO, pH and Temperature

# Convergent Cross Mapping

```{r}
n <- NROW(Phyto_std)
ccm_rho_matrix2 <- matrix(NA, nrow = length(vars), ncol = length(vars), dimnames = list(vars, 
    vars))

e = 1
for (ccm_from in vars) {
    for (ccm_to in vars[vars != ccm_from]) {
        out_temp <- ccm(Phyto_std, E = rimov.best_E[[e]], lib_column = ccm_from, target_column = ccm_to, 
            lib_sizes = n, replace = FALSE, silent = TRUE)
        ccm_rho_matrix2[ccm_from, ccm_to] <- out_temp$rho
    }
  e = e+1
}

ccm_rho_matrix2
```

## Acknowledgement

The analyses made below, were possible Two essential hat tips have to be made for the production of these analyses:

-   The rEDM tutorial

-   Empirical dynamic dynamic modelling analyses by Owen Liu. <file:///C:/Users/sdijoux/Dropbox/%23%2000.%20Research/Research%20projects/_2024_TS-Rimov/Analysis/env_context_dependency/edm_analysis.html#3_convergent_cross_mapping>
