---
title: "02.Rimov_EDM_analysis"
format:
  html:
    toc: true
    toc-depth: 4
    toc-title: Contents
    toc-location: left-body
    embed-resources: true
    self-contained-math: true
date: today
date-format: long
editor: visual
author: Samuel Dijoux
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

### Short overview

This script is dedicated to the empirical dynamical modelling analyses performed on the time series of morpho-functional groups of phytoplanktonic communities sampled in the Řimov reservoir (Czech Republic) between 1997 and 2021.

Each variable is sub

### Required materials

#### Libraries

Load the required packages (if these are not installed on your system, use `install.packages()`)

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(tidyr)
library(knitr)
library(kableExtra)
library(purrr)
library(dplyr)
library(tibble)
library(readr)
library(stringr)
library(lubridate)
library(forcats)
library(fields)

library(rEDM)
library(ncdf4)
library(quantreg)
library(RANN)
library(ggsci)
library(here)

library(ggplot2)
library(plot3D)
library(igraph)
library(gridExtra)
```

#### Plot theme

The following chunk defines the plot theme used across the analyses

```{r}
plot_theme <-   theme_minimal()+
  theme(text=element_text(family="sans",size=12,color="black"),
        legend.text = element_text(size=14),
        axis.title=element_text(family="sans",size=14),
        axis.text=element_text(family="sans",size=8),
        strip.background = element_rect(colour="black"),
        panel.border = element_rect(color="black",fill=NA))

theme_set(plot_theme)
```

#### Rimov dataset

```{r}
load("Data/Rimov.datasets.RData")
```

## Variables in the dataset

-   Description of the phytoplanktonic Morpho-Functional Groups (*MFGs*) variables

```{r}
tibble(Variable=colnames(Phyto_std),
       Description = c(paste0("Normalized biomass of MFG ", substr(colnames(Phyto_std), 5,  6) ) )) %>% knitr::kable()
```

-   Description of the environmental variables

```{r}
tibble(Variable=colnames(Env_std),
       Description = paste0("Normalized ", c(
  "Dissolved oxygen", "pH", "Conductometry", "Dissolved organic carbon",
  "Total Phosphorus concentration", "Total Nitrogen concentration",
  "Dissolved inorganic nitrogen", "Altitude of the water surface (m.a.s.l)", "Temperature" ) )) %>% knitr::kable()
```

## Establishing univariate predictability and nonlinearity

### Simplex projection and Embedding dimensions

we use Nearest neighbor forecasting method (Simplex) to determine the optimal embedding dimensions $E$ to produce forecasts for each MFG variables. This is done by selecting the $E$ for which we observe the highest forecast skill $\rho$.

```{r}
Rimov.phyto.simplex.list <- list()
for(i in 1:length(vars)){
  ts <- Phyto_std[[vars[i]]]
  Rimov.phyto.simplex.list[[i]] <- simplex(ts, lib=c(1, floor(dim(Phyto_std)[1]/2) ),
                              pred=c(floor(dim(Phyto_std)[1]/2)+1, dim(Phyto_std)[1]), silent=TRUE )
}
names(Rimov.phyto.simplex.list) <- vars
```

```{r}
par(mfrow = c(2, 2))
for (var in names(Rimov.phyto.simplex.list)) {
    plot(Rimov.phyto.simplex.list[[var]]$E, Rimov.phyto.simplex.list[[var]]$rho, type = "l", xlab = "Embedding Dimension (E)", 
        ylab = expression(paste("Skill, ",rho)), main = var)
}

# bind_rows(Rimov.phyto.simplex.list) %>%
#   ggplot(aes(E,rho))+
#   geom_line(size=2)+
#   #facet_wrap(nrow=2,scales="free_y")+
#   labs(x="Embedding Dimension (E)",y=expression(paste("Skill, ",rho)))+
#   scale_x_continuous(breaks=seq(0,12,by=2))
  
```

```{r}
# bind_rows(westend.simp.list) %>%
#   ggplot(aes(E,rho,color=spp))+
#   geom_line(size=2)+
#   facet_wrap(~spp,nrow=2,scales="free_y")+
#   labs(x="Embedding Dimension (E)",y=expression(paste("Skill, ",rho)))+
#   scale_x_continuous(breaks=seq(0,12,by=2))+
#   scale_color_manual(values=c('Macrocystis >1m'="navyblue",'Macrocystis <1m'="gray50",'Pterygophora'="darkgreen",'Laminaria'="darkcyan",'Purple urchin'="mediumpurple4",'Red urchin'="darkred"),name="")+
#   guides(color=F)
```

We saves the best dimension

```{r}
rimov.best_E <- sapply(Rimov.phyto.simplex.list, function(df) {
    df$E[which.max(df$rho)]
})
rimov.best_E
```

### S-maps of each species to identify nonlinear/linear time series

Using the optimal embedding dimensions previously analysed, we apply the S-map method to identify whether the variables follow a linear dynamic (if the highest predictive skill $\rho$ occurs for $\theta = 0$) or a nonlinear dynamic (higher $\rho$ observed for $\theta > 0$).

```{r}
rimov.phyto.smaps <- lapply(vars, function(var) {
    s_map(Phyto_std[[var]], E = rimov.best_E[var],
          lib=c(1, floor(dim(Phyto_std)[1]/2) ),
                              pred=c(floor(dim(Phyto_std)[1]/2)+1, dim(Phyto_std)[1])
          )
})
names(rimov.phyto.smaps) <- names(Rimov.phyto.simplex.list)
```

```{r}
par(mfrow = c(2, 2))
for (var in names(rimov.phyto.smaps)) {
    plot(rimov.phyto.smaps[[var]]$theta,
         rimov.phyto.smaps[[var]]$rho, type = "l", xlab = "Nonlinearity (theta)", 
        ylab = "Forecast Skill (rho)", main = var)
}
```

## Convergent Cross Mapping

```{r}
n <- NROW(Phyto_std)
ccm_rho_matrix2 <- matrix(NA, nrow = length(vars), ncol = length(vars), dimnames = list(vars, 
    vars))

e = 1
for (ccm_from in vars) {
    for (ccm_to in vars[vars != ccm_from]) {
        out_temp <- ccm(Phyto_std, E = rimov.best_E[[e]], lib_column = ccm_from, target_column = ccm_to, 
            lib_sizes = n, replace = FALSE, silent = TRUE)
        ccm_rho_matrix2[ccm_from, ccm_to] <- out_temp$rho
    }
  e = e+1
}

ccm_rho_matrix2
```

## Acknowledgement

The analyses made below, were possible Two essential hat tips have to be made for the production of these analyses:

-   The rEDM tutorial

-   Empirical dynamic dynamic modelling analyses by Owen Liu. <file:///C:/Users/sdijoux/Dropbox/%23%2000.%20Research/Research%20projects/_2024_TS-Rimov/Analysis/env_context_dependency/edm_analysis.html#3_convergent_cross_mapping>
